title: "Reverse Engineering Project"
author: "Alisha Camacho & Laura Cech"
date: "Date here"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction 

In this notebook, we are reverse engineering the story, [Thousands of kids are missing rom school. Where did they go?](https://projects.apnews.com/features/2023/missing-children/index.html)

## Load libraries

Loading required libraries for this analysis.

```{r echo=FALSE, message=FALSE}

library(tidyverse)
library(refinr)
library(janitor)
library(lubridate)
install.packages("readxl")
library(readxl)

```


### Sentence 2

* **Sentence text**: [Some of those students enrolled in private schools. Across the states in the analysis, private-school enrollment grew by about 103,000 students.]

* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

We filtered by the 21 states used in their analysis, and then found the sum of students enrolled in the school years "2021-22" and "2019-2020" and subtracted the difference. Our answer, 102,847 of students, is the same number, if we rounded to 103,000 students enrolled. 


```{r}

private_all21 <- private |> 
filter(state =="CA"| state=="NY" | state=="LA" | state=="NC"| state=="WA"| state=="CO" | state=="GA" | state=="MN" | state=="WI" | state=="NH" | state=="MA" | state=="MT" | state=="DE"| state=="ND" | state=="DC" | state=="SD" | state=="SC" | state=="NE" | state=="RI" | state=="TN" | state=="PA" | state=="FL") |> 
  group_by(sch_year) |> 
 # filter(sch_year=="2019-2020") |> 
  filter(grade=="k_12_total") |> 
  summarise(total=sum(enrollment))

glimpse(private_all21)

```

```{r}
private_2021_2022 <-private_all21 |> 
filter(sch_year == "2021-2022")

###
private_2019_2020 <- private_all21 |> 
filter(sch_year == "2019-2020")

###

private_2021_2022$total - private_2019_2020$total

```

### Sentence 3

* **Sentence text**: [ A much bigger group of students switched to home-schooling. The number of children registered for home-schooling surged by around 184,000.]


* **Analysis summary**: [In this instance, we converted the character values in the 'Total change' column  to numerical values. From there, we summarized the total change column to find a total of 184,047, which rounds down  to the answer of 184,000 students.]


```{r}
#Disclaimer, use of Chat CPT to create a .csv file from a tab in an excel sheet
#Prompt = "how do you create a .csv file from a tab in an excel sheet in Rstudio?"


missing_homeschool <- read_excel("/Users/AlishaCamacho/Documents/GitHub/Reverse_Engineering_Project/Data/Missing-Kids-Summary-File.xlsx", sheet = "Enrollment Changes")

write.csv(missing_homeschool, "missing_homeschool.csv", row.names = FALSE)

glimpse(missing_homeschool)

```


```{r}
homeschool_all21 <- missing_homeschool |> 
filter(`State abbrev` =="CA"| `State abbrev` =="NY" | `State abbrev`=="LA" | `State abbrev`=="NC"| `State abbrev`=="WA"| `State abbrev`=="CO" | `State abbrev`=="GA" | `State abbrev`=="MN" | `State abbrev`=="WI" | `State abbrev`=="NH" | `State abbrev`=="MA" | `State abbrev`=="MT" | `State abbrev`=="DE"| `State abbrev`=="ND" | `State abbrev`=="DC" | `State abbrev`=="SD" | `State abbrev`=="SC" | `State abbrev`=="NE" | `State abbrev`=="RI" | `State abbrev`=="TN" | `State abbrev`=="PA" | `State abbrev`=="FL") |> 
filter(`Enrollment type` == "Homeschool enrollment" )

glimpse(homeschool_all21)

```


```{r}

homeschool_all21 <- homeschool_all21 %>%
  mutate(
    sum_change = as.numeric(`Total change`)) |>
  summarise(sum_change = sum(sum_change, na.rm = TRUE))

glimpse(homeschool_all21)

```
 
