---
title: "Reverse Engineering Project"
author: "Alisha Camacho & Laura Cech"
date: "Date here"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction 

In this notebook, we are reverse engineering the story, [Thousands of kids are missing rom school. Where did they go?](https://projects.apnews.com/features/2023/missing-children/index.html)

## Load libraries

Loading required libraries for this analysis.

```{r echo=FALSE, message=FALSE}

library(tidyverse)
library(refinr)
library(janitor)
library(lubridate)
install.packages("readxl")
library(readxl)

```

## Load and Cleaning Data

In this section, describe the source of the data, write a basic data dictionary for data you are working with, and discuss any caveats or issues you discovered working with this data. 

```{r}
# Load required data

# Path to data should be loaded from folder "data" i.e. read_csv("data/name_of_data.csv")

# Clean required data and prepare for analysis if needed. 

population <- read_csv("/Users/AlishaCamacho/Documents/GitHub/Reverse_Engineering_Project/Data/population-by-age.csv")

private <- read_csv("/Users/AlishaCamacho/Documents/GitHub/Reverse_Engineering_Project/Data/private-enrollment-by-grade.csv")

public <- read_csv("/Users/AlishaCamacho/Documents/GitHub/Reverse_Engineering_Project/Data/public-enrollment-by-grade.csv")

missing <- read_excel("/Users/AlishaCamacho/Documents/GitHub/Reverse_Engineering_Project/Data/Missing-Kids-Summary-File.xlsx")



```
```{r}
#glimpse(population)
#glimpse(private)
#glimpse(public)
glimpse(missing)

```

## Sentences to Engineer
In this notebook, we are reverse engineering five sentences from the story.

### Sentence 1

* **Sentence text**: [Overall, public school enrollment fell by over 700,000 students between the 2019-2020 and 2021-2022 school years in the 21 states plus Washington, D.C., that provided the necessary data.]

* **Analysis summary**: [Write up two to three sentences describing the results of your analysis. Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

public_21 <- public |> 
filter(st =="CA"| st=="NY" | st=="LA" | st=="NC"| st=="WA"| st=="CO" | st=="GA" | st=="MN" | st=="WI" | st=="NH" | st=="MA" | st=="MT" | st=="DE"| st=="ND" | st=="DC" | st=="SD" | st=="SC" | st=="NE" | st=="RI" | st=="TN" | st=="PA" | st=="FL") |> 
  group_by(sch_year) |> 
 # filter(sch_year=="2019-2020") |> 
  filter(grade=="k_12_total") |> 
  summarise(total=sum(enrollment))
 
public_2019 <- public_21 |>
  filter(sch_year == "2019-2020")
public_2019_enrollment <- public_2019$total
public_2019_enrollment

###

public_2021 <- public_21 |> 
  filter(sch_year =="2021-2022")
public_2021_enrollment <-public_2021$total
public_2021_enrollment

public_2021_enrollment - public_2019_enrollment


# Display results of code below this codeblock
704593

```


p_2019 <- public |> 
#  group_by(state) |>   
  filter(sch_year == "2019-2020") |> 
  summarise(total=sum(enrollment, na.rm = TRUE))
  
p_2022 <- public |> 
#  group_by(state) |>   
  filter(sch_year == "2021-2022") |> 
  summarise(total=sum(enrollment, na.rm = TRUE))
  
```

### Sentence 2

* **Sentence text**: [Some of those students enrolled in private schools. Across the states in the analysis, private-school enrollment grew by about 103,000 students.]

* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]


```{r}
# Put code to reverse engineer sentence here

glimpse(private)

# Display results of code below this codeblock

```

### Sentence 3

* **Sentence text**: [ A much bigger group of students switched to home-schooling. The number of children registered for home-schooling surged by around 184,000.]

* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

### Sentence 4

* **Sentence text**: [Paste in sentence to engineer here]
* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

### Sentence 5

* **Sentence text**: [Paste in sentence to engineer here]
* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

-30-